Using Monte Carlo Algorithms to Achieve Resiliency \& Performance at Scale for Linear and Nonlinear Solver Applications

### 2016-01-26 doc/ecp/pres.tex
Updated presentation.

\documentclass{beamer}
%% The amssymb package provides various useful mathematical symbols
\usepackage{amssymb}
%% The amsthm package provides extended theorem environments
\usepackage{amsthm} \usepackage{amsmath} \usepackage{tmadd,tmath}
\usepackage{booktabs}
\usepackage{multirow}
\usepackage[compatibility=false]{caption}
\usepackage[labelformat=empty]{subcaption}
\usepackage{tikz}
\usepackage{ragged2e}

@@

  \begin{columns}
    \begin{column}{0.5\textwidth}
      \begin{itemize}
      \item Each parallel process owns a piece of the domain (linear
        system)
        \bigskip
      \item Random walks must be transported between adjacent domains
        through parallel communication
        \bigskip
      \item Domain decomposition determined by the input system
        \bigskip
      \item Load balancing not yet addressed
        \bigskip
      \item Uses modified version of the Milagro algorithm (LANL)
      \end{itemize}
    \end{column}

@@

\begin{frame}{Remaining Research}

@@

  \underline{Presentations}
  \begin{itemize}
  \item Monte Carlo Synthetic Acceleration Methods for Sparse Linear Systems
    (M.L. Pasini, M. Benzi, T. Evans, S. Hamilton, S. Slattery), \textit{SIAM
      ALA}, October 30, 2015.
  \item Parallel Algorithms for the Monte Carlo Synthetic Acceleration Linear
    Solver Method (S. Slattery, T. Evans, S. Hamilton), \textit{SIAM CSE},
    Salt Lake City, 2015.
  \item Iterative Performance of Monte Carlo Linear Solver Methods
    (M.L. Pasini), \textit{SIAM CSE}, Salt Lake City, 2015.
  \item A multilevel Monte Carlo method for linear systems (S. Slattery,
    T. Evans, S. Hamilton), \textit{Copper Mountain Conference on Iterative
      Solvers}, Copper MT, CO, 2014.
  \item Monte Carlo Synthetic Acceleration as approximate polynomial
    preconditioning (S. Hamilton, T. Evans, S. Slattery), \textit{Copper
      Mountain Conference on Iterative Solvers}, Copper MT, CO, 2014.
  \item Monte Carlo Linear Solvers (S. Slattery, T. Evans, S. Hamilton),
    \textit{NC State Seminar}, Nov. 4, 2014.
  \item Monte Carlo Linear Solvers (S. Hamilton, T. Evans, S. Slattery),
    \textit{Emory Mathematics Dept. Seminar}, Sept. 19, 2014.
  \item Monte Carlo Methods for Linear Systems (T. Evans), \textit{MIT Nuclear
      Engineering Dept Seminar Series}, Sept. 30, 2014.
  \end{itemize}

@@

@@

  MC methods are used in many DOE application areas:
  \begin{itemize}
  \item high-energy density physics
  \item astrophysics
  \item nuclear engineering
  \item particle accelerators and neutron scattering
  \item quantum chemistry
  \item visualization
  \item fluid dynamics
  \item computational biology
  \end{itemize}

@@

\begin{frame}{Continued Research for MC Linear Solvers}

@@

@@

  MCREX has leveraged and built on work derived from several ASCR projects:
  \begin{itemize}
  \item Trilinos
  \item XPress
  \item xSim
  \item Ideas
  \end{itemize}

>>>>

\documentclass{beamer}
%% The amssymb package provides various useful mathematical symbols
\usepackage{amssymb}
%% The amsthm package provides extended theorem environments
\usepackage{amsthm} \usepackage{amsmath} \usepackage{tmadd,tmath}
\usepackage{booktabs}
\usepackage{multirow}
\usepackage{subfigure}
\usepackage[compatibility=false]{caption}
\usepackage{tikz}
\usepackage{ragged2e}
\usepackage{algorithm,algorithmic}

@@

  \begin{columns}
    \begin{column}{0.5\textwidth}
      \begin{itemize}
        \setlength{\itemsep}{-0.4\baselineskip}
      \item Each parallel process owns a piece of the domain (linear
        system)
        \bigskip
      \item Random walks must be transported between adjacent domains
        through parallel communication
        \bigskip
      \item Domain decomposition determined by the input system
        \bigskip
      \item Load balancing not yet addressed
        \bigskip
      \item Uses modified version of the Milagro algorithm (LANL)
      \end{itemize}
    \end{column}

@@

\begin{frame}{Continued Research is Necessary}

@@

  \underline{Presentations}
  \begin{itemize}
    \footnotesize
    \setlength{\itemsep}{-0.1\baselineskip}
  \item Monte Carlo Synthetic Acceleration Methods for Sparse Linear Systems
    (M.L. Pasini, M. Benzi, T. Evans, S. Hamilton, S. Slattery), \textit{SIAM
      ALA}, October 30, 2015.
  \item Parallel Algorithms for the Monte Carlo Synthetic Acceleration Linear
    Solver Method (S. Slattery, T. Evans, S. Hamilton), \textit{SIAM CSE},
    Salt Lake City, 2015.
  \item Iterative Performance of Monte Carlo Linear Solver Methods
    (M.L. Pasini), \textit{SIAM CSE}, Salt Lake City, 2015.
  \item A multilevel Monte Carlo method for linear systems (S. Slattery,
    T. Evans, S. Hamilton), \textit{Copper Mountain Conference on Iterative
      Solvers}, Copper MT, CO, 2014.
  \item Monte Carlo Synthetic Acceleration as approximate polynomial
    preconditioning (S. Hamilton, T. Evans, S. Slattery), \textit{Copper
      Mountain Conference on Iterative Solvers}, Copper MT, CO, 2014.
  \item Monte Carlo Linear Solvers (S. Slattery, T. Evans, S. Hamilton),
    \textit{NC State Seminar}, Nov. 4, 2014.
  \item Monte Carlo Linear Solvers (S. Hamilton, T. Evans, S. Slattery),
    \textit{Emory Mathematics Dept. Seminar}, Sept. 19, 2014.
  \item Monte Carlo Methods for Linear Systems (T. Evans), \textit{MIT Nuclear
      Engineering Dept Seminar Series}, Sept. 30, 2014.
  \end{itemize}

@@

  Using the MCLS interface to Trilinos, it is straightforward to implement
  MCSA in other applications
  \vfill
  We implemented MCSA in the production CFD code, Drekar (SNL) using a
  Newton-MCSA algorithm:
  \begin{algorithm}[H]
    \footnotesize
    \begin{algorithmic}[1]
      \STATE $k := 0$
      \WHILE{$||\ve{F}(\ve{u}^{k})|| > \epsilon
        ||\ve{F}(\ve{u}^{0})||$}
      \STATE $\ve{J}(\ve{u}^{k}) \leftarrow AD(\ve{F}(\ve{u}^k))$
      \COMMENT{Automatic differentiation - Panzer}
      \STATE $\ve{J}(\ve{u}^k) \delta \ve{u}^k = -\ve{F}(\ve{u}^{k})$
      \COMMENT{Solve for the Newton correction with MCSA}
      \STATE $\ve{u}^{k+1} \leftarrow \ve{u}^k + \delta \ve{u}^k$
      \STATE $k \leftarrow k+1$
      \ENDWHILE
    \end{algorithmic}
    \caption{FANM}
  \end{algorithm}
  \vfill
  Because MCLS was built with Trilinos abstract numerical algorithms this
  integration was straightforward and we can test MCLS in a whole new problem
  space
  \vfill
  We are also able to utilize all of Trilinos' preconditioners

\begin{frame}{Thermal Convection Cavity Results}

  \begin{columns}
    \begin{column}{0.6\textwidth}

      \small{
        \begin{table}[h!]
          \begin{center}
            \begin{tabular}{lccc}\hline\hline
              \multicolumn{1}{l}{Benchmark}&
              \multicolumn{1}{c}{NK}&
              \multicolumn{1}{c}{FANM}&
              \multicolumn{1}{c}{NR}\\
              \hline
              Ra=\sn{1}{3} & 5 & 5 & 5\\
              Ra=\sn{1}{4} & 7 & 7 & 7\\
              Ra=\sn{1}{5} & 9 & 10 & 9\\
              Ra=\sn{1}{6} & 11 & 11 & 11\\
              %%
              \hline\hline
            \end{tabular}`
          \end{center}
          \caption{Nonlinear iterations.}
        \end{table}

        \begin{table}[h!]
          \begin{center}
            \begin{tabular}{lccc}\hline\hline
              \multicolumn{1}{l}{Benchmark}&
              \multicolumn{1}{c}{GMRES}&
              \multicolumn{1}{c}{MCSA}&
              \multicolumn{1}{c}{Richardson}\\
              \hline
              Ra=\sn{1}{3} & 32 & 18 & 38\\
              Ra=\sn{1}{4} & 23 & 17 & 34\\
              Ra=\sn{1}{5} & 25 & 20 & 34\\
              Ra=\sn{1}{6} & 39 & 25 & 48\\
              %%
              \hline\hline
            \end{tabular}
          \end{center}
          \caption{Total linear solver iterations.}
        \end{table}
      }

    \begin{column}{0.4\textwidth}

      \small{
        \begin{itemize}
        \item Weaker preconditioning at Ra=\sn{1}{5} case
          \medskip
        \item Constant forcing term for Ra=\sn{1}{6} case
        \end{itemize}
      }

      \bigskip

      \tiny{
        \begin{table}[h!]
          \begin{center}
            \begin{tabular}{lc}\hline\hline
              \multicolumn{1}{l}{Benchmark}&
              \multicolumn{1}{c}{NK Speedup}\\
              \hline
              Ra=\sn{1}{3} & 338 \\
              Ra=\sn{1}{4} & 336 \\
              Ra=\sn{1}{5} & 346 \\
              Ra=\sn{1}{6} & 465 \\
              \hline\hline
            \end{tabular}
          \end{center}
          \caption{Newton-Krylov speedup over FANM.}
        \end{table}
      }

    \end{column}
  \end{columns}

@@

  MC methods are used in many DOE application areas:
  \begin{itemize}
    \setlength{\itemsep}{-0.2\baselineskip}
  \item high-energy density physics
  \item astrophysics
  \item nuclear engineering
  \item particle accelerators and neutron spallation source
  \item quantum chemistry
  \item visualization
  \item fluid dynamics
  \item combustion
  \item computational biology
  \end{itemize}

@@

\begin{frame}{ECP R\&D I: Time-dependent Problems}

@@

\begin{frame}{ECP R\&D II: Monte Carlo Particle Transport}

  MC particle transport has applications in many areas across the DOE complex
  \begin{itemize}
    \setlength{\itemsep}{-0.2\baselineskip}
  \item high-energy density physics
  \item neutron spallation source
  \item nuclear engineering
  \item astrophysics
  \item combustion
  \end{itemize}

  There are significant challenges in porting random walk algorithms to modern
  heterogeneous computing devices
  \begin{enumerate}
  \item Random walks are dominated by branching statements $\rightarrow$
    thread divergence
  \item Runtime is often dominated by sampling and interpolating large fields
    of data (keeping the GPUs fed)
  \end{enumerate}

  Supporting research into the efficient parallelization and instrumentation
  of random walk algorithms for particle transport will benefit many
  applications in DOE

\begin{frame}{ECP R\&D III: Particle Methods}

  Non-stochastic particle methods share a great deal of algorithmic features
  with stochastic particle methods
  \begin{itemize}
  \item Particles must be ordered and sorted efficiently
  \item Particles must be moved through a combinatorial and mesh-based
    geometries
  \item Data structures must be optimized for efficient vectorization (over
    32-thread warps on GPUs)
  \item Inter-node communication of particle data structures are required for
    domain decomposition parallel topologies
  \item Physical parameters must be mapped between particles and an underlying
    data representation (grid)
  \end{itemize}

  A supporting technolgies research effort on a generalized set of libraries
  for managing particle-based data structures on heterogeneous computing
  platforms will benefit many applications in DOE

@@

  \underline{\bf Trilinos}
  \begin{itemize}
  \item Provides abstract solver interface for MCLS
  \item Provides build system for Alea and MCLS
  \item Provides data structures and utilities for Alea and MCLS
  \end{itemize}

  \underline{\bf xSIM}
  \begin{itemize}
  \item Simulator to test for hard error detection and remediation
  \end{itemize}

  \underline{\bf XPRESS}
  \begin{itemize}
  \item We are currently implementing a CUDA transport kernel with scheduling
    optimized by HPX as part of XPRESS
  \item We have already compared OpenMP versus HPX for on-node MC particle
    transport algorithms
  \end{itemize}

  \begin{figure}
    \centering
    \begin{tabular}{cc}
      \subfigure[time]{\includegraphics[width=2in]{time}}
      &
      \subfigure[efficiency]{\includegraphics[width=2in]{tallies}}
    \end{tabular}
  \end{figure}
